---
title: "Project proposal"
author: "Tutorial 3 Team 10"
output: github_document
---





## 1. Introduction
Whilst looking for a dataset for our project, we came across one about suicides. This got us wondering about how suicide rates are influenced by different variables. For our project we will be looking into how suicides differ between genders, countries, and each year from 1985 to 2016. We will also compare the countries to other variables such as HDI and gdp per capita, in order to extract valuable new information of what affects the rates of suicide in each country.

Hypothesis:
We suspect that countries with low gdp per capita and low Human Development Index will have high levels of suicide. We also think that men will have higher suicide rates than women.

The data was collected from kaggle and uses data from the United Nations Development Program. (2018). Human development index (HDI). Retrieved from http://hdr.undp.org/en/indicators/137506; World Bank. (2018). World development indicators: GDP (current US$) by country:1985 to 2016. Retrieved from http://databank.worldbank.org/data/source/world-development-indicators# ; [Szamil]. (2017). Suicide in the Twenty-First Century [dataset]. Retrieved from https://www.kaggle.com/szamil/suicide-in-the-twenty-first-century/notebook ; World Health Organization. (2018). Suicide prevention. Retrieved from http://www.who.int/mental_health/suicide-prevention/en/. 
The observations are the number of suicides that fall into each category of gender, age group, country and year. There are in total 27,280 observations spanning 101 countries and 31 years. The variables are: year, sex, age, number of suicides, population, number of suicides per one hundred thousand people, the HDI (Human Development Index) of each country, the GDP of each country, the GDP per capita of each country and the generation of the victims.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readr)
library(forcats)
suicides <- read_csv("master.csv")
view(suicides)
```

## 2. Data

```{r display-dataframe, message = FALSE}
suicides <- read_csv("master.csv")
glimpse(suicides)
```

## 3. Data analysis plan

```{r renaming}
suicides <- suicides %>% 
  rename(hdi = 'HDI for year',
         suicides_per_100k_pop = 'suicides/100k pop',
         country_year = 'country-year',
         gdp ='gdp_for_year ($)',
         gdp_per_capita = 'gdp_per_capita ($)') 
```


```{r total-suicides-country}
suicides %>% 
  group_by(country) %>% 
  summarise(n = sum(suicides_no)) %>% 
  arrange(desc(n))
```

First we summed the total suicides by country to see which country had the greatest number of suicides in the dataset.This turned out to be the Russian Federation. However this does not tell us much about suicide rates as it doesn't take into account the population of the country. In the next code chunk we will sum the total number of suicides per 100000 population to see which country has the highest suicide rates.

```{r total-suicides-country-per-pop}
suicides %>% 
  group_by(country) %>% 
  summarise(n = sum(suicides_per_100k_pop)) %>% 
  arrange(desc(n))
```

It seems that Russia also has the highest total number of suicides per 100000 people in the 30 years. However the next highest is Lithuania which has a small population but a high level of suicide.

In our next plot we test one of our hypotheses. Do more men commit suicide than women?

```{r suicides_by_gender}
suicides %>% 
  group_by(sex) %>% 
  summarise(n = sum(suicides_no)) %>% 
  ggplot(aes(x = sex, y = n)) + 
  geom_col() +
  labs(title = "Suicide Rates by Gender",
       x = "Gender",
       y = "Number of Cases")
```

This graph shows our hypothesis is correct, there is almost three times more male suicides than female.

Our final plot compares the total number of suicides for each year.

```{r suicides_by_year}
suicides %>% 
  group_by(year) %>% 
  summarise( n = sum(suicides_no)) %>% 
  ggplot(aes(x = year, y = n)) + 
  geom_line() +
  labs(title = "Suicide Rates by Year",
       x = "Year",
       y = "Number of Cases")
```

The value for 2016 is extremely low. This suggests that the data for 2016 is incomplete so we should ignore this. Suicide rates were highest around the late 1990s, early 2000s. There is also a increase around 2008/ 2009. One possible contributor to this is the 2008 financial crash.
