---
title: "The Rate of Suicides by Various Factors"
author: "Tutorial 3 Team 10 <br> Michael Beer, Ronan Blackwood, Fernando Gomes and Ed Grosvenor"
institute: "University of Edinburgh"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(readr)
library("ggplot2")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library(rgeos)
theme_set(theme_void())
suicides <- read_csv("../data/master.csv")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

#The data set

<ul>
<br>
--
<font size="13"><li>This data set records the number of suicides per year for each country.<br><br></font>

--
<font size="13"><li>There are in total 27,280 observations spanning 101 countries and 31 years.<br><br></font>

--
<font size="13"><li>The data was collected from Kaggle.<br><br></font>
</ul>

---

```{r suicide, echo = FALSE, include=FALSE}
suicides_clean <- suicides %>% 
  rename(hdi = 'HDI for year',
         suicides_per_100k_pop = 'suicides/100k pop',
         country_year = 'country-year',
         gdp ='gdp_for_year ($)',
         gdp_per_capita = 'gdp_per_capita ($)') 
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r suicide-average, echo = FALSE, include=FALSE}
suicide_rates <- suicides_clean %>%
  group_by(country) %>% 
  summarise(suicide_rates = mean(suicides_per_100k_pop)) %>% 
  arrange(desc(suicide_rates))
```

```{r world, echo = FALSE, include=FALSE}
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(type == "Sovereign country" | type == "Country")
world_points <- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))
```

```{r total-suicides-country-per-pop, echo = FALSE, include=FALSE}
  suicide_heat <- suicide_rates %>%
  rename(name_long = 'country')
```

```{r world-map-suicides, echo = FALSE, include=FALSE}
  suicidemap <- merge(world, suicide_heat, by="name_long", all=TRUE)
```

#Heat map of the world
--
```{r heatmap, echo = FALSE, out.width = "100%"}
  worldheat <- ggplot(data = suicidemap) +
    geom_sf(aes(fill = suicide_rates)) +
        labs(title = "Suicide Rates Per Country", fill = "Suicide Rates Per \n 100k People") +
           scale_fill_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="grey") +
                            theme(panel.background = element_rect(fill = "ivory", color = "black"))
  worldheat
```
class: middle, center
---

#Why a heat map?
<ul>
<br>
--
<font size="13"><li>Great way of visualizing the data.<br><br></font>

--
<font size="13"><li>Clean and easy to understand.<br><br></font>

--
<font size="13"><li>You can fit all the countries on this plot.<br><br></font>
</ul>

---

#We can zoom into Europe

.pull-left[
```{r europe, echo = FALSE, out.width= "100%"}
worldheat +
  coord_sf(xlim = c(-30, 60), ylim = c(30, 80), expand = FALSE) +
  labs(title = "Suicide Rates Per Country (Europe)")
```
]
--
.pull-right[
<ul>
<br><font size="6"><li>Europe can be seen to have relatively high suicide rates overall.</font>
<br><br>
--
<font size="6"><li>Especially in for countries in Eastern Europe.</font>
</ul>
]
---

#What about South America?
--
.pull-left[
```{r southamerica, echo = FALSE, out.width = "100%"}
worldheat +
  coord_sf(xlim = c(-90, -30), ylim = c(-60, 15), expand = FALSE) +
      labs(title = "Suicide Rates Per Country (South America)")
```
]
--
.pull-right[
```{r southamericatxt, echo = FALSE, out.width = "100%"}
worldheat +
  coord_sf(xlim = c(-90, -30), ylim = c(-60, 15), expand = FALSE) +
  geom_text(data= world_points,aes(x=X, y=Y, label=name),
    color = "black", check_overlap = TRUE) +
      labs(title = "Suicide Rates Per Country (South America) - annotated")
```
]
--
.footnote[
<font size="6">South America can be seen to have lower suicide rates overall.</font>
]

---

#What did we learn?

<ul>
<br>
--
<font size="13"><li>Russia has the worst suicide rates for a major country.<br><br></font>

--
<font size="13"><li>Eastern Europe seem to have high suicide rates.<br><br></font>

--
<font size="13"><li>There is a lot of missing countries in the data set.<br><br></font>
</ul>

---

#Gender

```{r male-suicides, echo = FALSE, include=FALSE}
male_suicides <- suicides_clean %>%
  filter(sex == "male") %>%
  group_by(year) %>%
  summarise(suicide_count = mean(suicides_no), sex = "male") %>%
  arrange(desc(suicide_count))
```

```{r fem-suicides, echo = FALSE, include=FALSE}
fem_suicides <- suicides_clean %>%
  filter(sex == "female") %>%
  group_by(year) %>%
  summarise(suicide_count = mean(suicides_no), sex = "female") %>%
  arrange(desc(suicide_count))
  
```
--
.pull-left[
```{r fem-table, echo=FALSE}
kable(head(fem_suicides), format = "html")
```
]
--
.pull-right[
```{r male-table, echo=FALSE}
kable(head(male_suicides), format = "html")
```
]
--
.footnote[
<font size="6">The male suicide count can be seen as alot higher than females.</font>
]
---
#Gender graph plot

```{r gen-suicides, echo=FALSE, include=FALSE}
gen_suicides <- rbind(male_suicides, fem_suicides)
```
--
.pull-left[
```{r suicide_by_gender_plot, echo=FALSE}
ggplot(gen_suicides,aes(x = year, y = suicide_count)) +
  geom_point(pch = 3) +
  geom_smooth(aes(color = sex), method = 'loess', formula = y ~ x) +
  labs(title = "Number of Suicides by Gender", x = "Year", y ="Number of Suicides", colour = "Sex") +
  scale_y_continuous(breaks = c(50000, 100000, 150000, 200000), labels = c("50k", "100k", "150k", "200k"))
```
]
.pull-right[
<ul>
<br><font size="4"><li>T</font>
<br><br>
--
<font size="4">There is not much of a trend between HDI and suicide rates.<li>.</font>
</ul>
]
---
#Human Development Index

--
<font size="6">The Human Development Index (HDI) is statistic calculated from life expectancy, education quality etc.</font>
--

```{r suicide-average-hdi, echo = FALSE, warning=FALSE, include=FALSE}
suicide_rates_hdi <- suicides_clean %>%
  filter(!is.na(hdi)) %>%
  group_by(country) %>%
  summarise(suicide_rates = round(mean(suicides_per_100k_pop),digits = 1), mean_hdi = round(mean(hdi),digits = 2)) %>% 
  arrange(desc(suicide_rates))
```
<br><br>
```{r hdi-table, echo=FALSE}
kable(head(suicide_rates_hdi), format = "html")
```
---

#HDI Plot
--
.pull-left[
```{r hdi-graph, echo=FALSE, warning= FALSE, out.width="60%"}
suicide_rates_hdi %>%
  ggplot(aes(x = mean_hdi, y = suicide_rates)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "red", formula = y ~ x) +
  labs(title = "Rate of Suicides Against HDI",
       x = "Human Development Index",
       y = "Suicides per 100k Population")
``` 
]
--
.pull-right[
<ul>
<br><font size="4"><li>The smooth line is using the formula 'y ~ x'.</font>
<br><br>
--
<font size="4">There is not much of a trend between HDI and suicide rates.<li>.</font>
</ul>
]
---

#What did we learn?

<ul>
<br>
--
<font size="13"><li>There seems to be no correlation between HDI and suicide rates.<br><br></font>

--
<font size="13"><li>The lowest country HDI on our data set was Guatemala with 0.58.<br><br></font>

--
<font size="13"><li>This shows most of the countries included in the ds had a relatively high HDI.<br><br></font>
</ul>

---

#Age of victim

```{r age-group, echo=FALSE, include=FALSE}
suicides_generation <- suicides_clean %>%
  group_by(age) %>%
  summarise(suicides = sum(suicides_no)) %>%
  arrange(age)
```

```{r age-plot, echo=FALSE}
suicides_generation %>%
  ggplot(aes(x = age, y = suicides)) +
  geom_col(fill = "lightblue") +
  labs(y = "Suicides", x = "Age") +
  scale_y_continuous(breaks = c(500000, 1000000, 1500000, 2000000, 2500000), labels = c("50k", "100k", "150k", "200k", "250k"))
```


