---
title: "The Rate of Suicides by Various Factors"
author: "Tutorial 3 Team 10 <br> Michael Beer, Ronan Blackwood, Fernando Gomes and Ed Grosvenor"
institute: "University of Edinburgh"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(readr)
library("ggplot2")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library(rgeos)
suicides <- read_csv("../data/master.csv")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

#The data set

<ul>
<br>
--
<font size="13"><li>This data set records the number of suicides per year for each country.<br><br></font>

--
<font size="13"><li>There are in total 27,280 observations spanning 101 countries and 31 years.<br><br></font>

--
<font size="13"><li>The data was collected from Kaggle.<br><br></font>
</ul>

---

```{r suicide, echo = FALSE, include=FALSE}
suicides_clean <- suicides %>% 
  rename(hdi = 'HDI for year',
         suicides_per_100k_pop = 'suicides/100k pop',
         country_year = 'country-year',
         gdp ='gdp_for_year ($)',
         gdp_per_capita = 'gdp_per_capita ($)') 
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r suicide-average, echo = FALSE, include=FALSE}
suicide_rates <- suicides_clean %>%
  group_by(country) %>% 
  summarise(suicide_rates = mean(suicides_per_100k_pop)) %>% 
  arrange(desc(suicide_rates))
```

```{r world, echo = FALSE, include=FALSE}
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(type == "Sovereign country" | type == "Country")
world_points <- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))
```

```{r total-suicides-country-per-pop, echo = FALSE, include=FALSE}
  suicide_heat <- suicide_rates %>%
  rename(name_long = 'country')
```

```{r world-map-suicides, echo = FALSE, include=FALSE}
  suicidemap <- merge(world, suicide_heat, by="name_long", all=TRUE)
```

#Heat map of the world
--
```{r heatmap, echo = FALSE, out.width = "100%"}
  worldheat <- ggplot(data = suicidemap) +
    geom_sf(aes(fill = suicide_rates)) +
        labs(title = "Suicide Rates Per Country", fill = "Suicide Rates Per \n 100k People") +
           scale_fill_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="grey") +
                            theme(panel.background = element_rect(fill = "ivory", color = "black"))
  worldheat
```
class: middle, center
---

#Why a heat map?
<ul>
<br>
--
<font size="13"><li>Great way of visualizing the data.<br><br></font>

--
<font size="13"><li>Clean and easy to understand.<br><br></font>

--
<font size="13"><li>You can fit all the countries on this plot.<br><br></font>
</ul>

---

#We can zoom into Europe

.pull-left[
```{r europe, echo = FALSE, out.width= "100%"}
worldheat +
  coord_sf(xlim = c(-30, 60), ylim = c(30, 80), expand = FALSE) +
  labs(title = "Suicide Rates Per Country (Europe)")
```
]
--
.pull-right[
<ul>
<br><font size="6"><li>Europe can be seen to have relatively high suicide rates overall.</font>
<br><br>
<font size="6"><li>Especially in for countries in Eastern Europe.</font>
</ul>
]
---

#What about South America?
--
.pull-left[
```{r southamerica, echo = FALSE, out.width = "100%"}
worldheat +
  coord_sf(xlim = c(-90, -30), ylim = c(-60, 15), expand = FALSE) +
      labs(title = "Suicide Rates Per Country (South America)")
```
]
--
.pull-right[
```{r southamericatxt, echo = FALSE, out.width = "100%"}
worldheat +
  coord_sf(xlim = c(-90, -30), ylim = c(-60, 15), expand = FALSE) +
  geom_text(data= world_points,aes(x=X, y=Y, label=name),
    color = "black", check_overlap = TRUE) +
      labs(title = "Suicide Rates Per Country (South America) - annotated")
```
]
--
.footnote[
<font size="6">South America can be seen to have lower suicide rates overall.</font>
]

---

#What did we learn?

<ul>
<br>
--
<font size="13"><li>Russia has the worst suicide rates for a major country.<br><br></font>

--
<font size="13"><li>Eastern Europe seem to have high suicide rates.<br><br></font>

--
<font size="13"><li>There is a lot of missing countries in the data set.<br><br></font>
</ul>

---

#Whats wrong with this graph?

--

.pull-left[
<font size="8">Comment in the chat why you think there is a massive drop in suicides.</font>
]
--
.pull-right[
```{r First plot, echo=FALSE, include=FALSE, out.width="60%"}
suicide_years <- suicides_clean %>% 
  group_by(year) %>% 
  summarise(n = sum(suicides_no))
```

```{r wrong-graph, echo=FALSE}
ggplot(data= suicide_years, mapping = aes(x = year, y = n)) +
geom_line() +
labs(title = "Total Number of Suicides per Year",
      x = "Year",
       y = "Total Number of Suicides"
       )
```
]
--

.footnote[
<font size="8">Not all countries collected data for the early and late years, especially 2016.<br><br></font>
]

---

#What about gender?

```{r male-suicides, echo = FALSE, include=FALSE}
male_suicides <- suicides_clean %>%
  filter(sex == "male") %>%
  group_by(year) %>%
  summarise(suicide_mean = mean(suicides_per_100k_pop), sex = "male") %>%
  arrange(desc(suicide_mean))
```

```{r fem-suicides, echo = FALSE, include=FALSE}
fem_suicides <- suicides_clean %>%
  filter(sex == "female") %>%
  group_by(year) %>%
  summarise(suicide_mean = mean(suicides_per_100k_pop), sex = "female") %>%
  arrange(desc(suicide_mean))
  
```
--
.pull-left[
```{r fem-table, echo=FALSE}
kable(head(fem_suicides), format = "html")
```
]
--
.pull-right[
```{r male-table, echo=FALSE}
kable(head(male_suicides), format = "html")
```
]
--
.footnote[
<font size="6">The male suicide count can be seen as alot higher than females.</font>
]
---
#Gender graph plot

```{r gen-suicides, echo=FALSE, include=FALSE}
gen_suicides <- rbind(male_suicides, fem_suicides)
```
--
.pull-left[
```{r suicide_by_gender_plot, echo=FALSE}
ggplot(gen_suicides,aes(x = year, y = suicide_mean)) +
  geom_point(pch = 3) +
  geom_smooth(aes(color = sex), method = 'loess', formula = y ~ x) +
  labs(title = "Number of Suicides by Gender", x = "Year", y ="Mean suicides per 100k Population", colour = "Sex")
```
]
--
.pull-right[
<font size="8">This clearly supports the statistic that male suicide is more frequent than female.</font>

]

---
#What did we learn?

<ul>
<br>
--
<font size="13"><li>Male suicide is much more common than female.<br><br></font>

--
<font size="13"><li>Male suicides peaked around 1995.<br><br></font>

--
<font size="13"><li>Female suicides are quite stable with no peaks<br><br></font>
</ul>

---

#More on the year variable
--

.pull-left[
<font size="13">Type here<br><br></font>
]
--
.pull-right[
```{r east-european-year, echo=FALSE, include=FALSE}
east_europe_suicides = suicides_clean %>% 
  filter(country %in% c("Russian Federation", "Lithuania", "Hungary", "Belarus", "Latvia", "Estonia", "Ukraine", "Slovenia")) %>% 
  group_by(year, country) %>% 
  summarise(total_suicides = mean(suicides_per_100k_pop))
```

```{r east-european-year-plot, echo=FALSE}
  ggplot(data = east_europe_suicides, aes(x = year, y = total_suicides, color = country)) + 
  geom_line() + 
  facet_wrap(~ country, ncol = 4) +
  labs(title = "Suicide Rates per Year in Eastern European Countries",
       x = "Year",
       y = "Total number of suicides") +
  guides(color = FALSE) 
```
]
---
#What did we learn?

<ul>
<br>
--
<font size="13"><li>Between 1995 and 2015 there is a steady decrease in  total number of suicides for these 8 countries. There are 4 countries in this set that have data for 1990, (Belarus, Hungary, Russia and Ukraine.) Belarus, Russia and Ukraine have an increase from 1990 to 1995. It is interesting to note that these are former Soviet Union countries which broke up in 1991. This may have had an effect on the suicide rates but there are many factors at play that it is hard to come to a definite conclusion. <br><br></font>

--
<font size="13"><li>Another hypothesis is that these trends are linked to HDI. Data from knoema:
https://knoema.com/atlas/Ukraine/Human-development-index?compareTo=RU,BY,HU shows that for Russia and Ukraine there is a dip in HDI around 1995. This gives us another hypothesis that there is a negative relationship between HDI and average suicide rates. <br><br></font>

--
<font size="13"><li>We will use a linear model to investigate this relationship. <br><br></font>
</ul>

---

#Human Development Index

--
<font size="6">The Human Development Index (HDI) is statistic calculated from life expectancy, education quality etc.</font>
--

```{r suicide-average-hdi, echo = FALSE, warning=FALSE, include=FALSE}
suicide_rates_hdi <- suicides_clean %>%
  filter(!is.na(hdi)) %>%
  group_by(country) %>%
  summarise(suicide_rates = round(mean(suicides_per_100k_pop),digits = 1), mean_hdi = round(mean(hdi),digits = 2)) %>% 
  arrange(desc(suicide_rates))
```
<br><br>
```{r hdi-table, echo=FALSE}
kable(head(suicide_rates_hdi), format = "html")
```
---

#HDI Plot
--
.pull-left[
```{r hdi-graph, echo=FALSE, warning= FALSE, out.width="100%"}
suicide_rates_hdi %>%
  ggplot(aes(x = mean_hdi, y = suicide_rates)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "red", formula = y ~ x) +
  labs(title = "Rate of Suicides Against HDI",
       x = "Human Development Index",
       y = "Suicides per 100k Population")
``` 
]
--
.pull-right[
<ul>
<br><font size="7"><li>The smooth line is using the formula 'y ~ x'.</font>
<br>
--
<font size="7">There is not much of a trend between HDI and suicide rates.<li>.</font>
</ul>
]
---

#What did we learn?

<ul>
<br>
--
<font size="13"><li>There seems to be no correlation between HDI and suicide rates.<br><br></font>

--
<font size="13"><li>The lowest country HDI on our data set was Guatemala with 0.58.<br><br></font>

--
<font size="13"><li>This shows most of the countries included in the ds had a relatively high HDI.<br><br></font>
</ul>

---

#Age of victim

--
```{r age-group, echo=FALSE, include=FALSE}
suicides_generation <- suicides_clean %>%
  group_by(age) %>%
  summarise(suicides = sum(suicides_no)) %>%
  slice(match(c("5-14 years","15-24 years","25-34 years","35-54 years","55-74 years","75+ years"), age))
```

```{r age-plot, echo=FALSE}
suicides_generation %>%
  mutate(age = fct_relevel(age, 
            "5-14 years", "15-24 years", "25-34 years", 
            "35-54 years", "55-74 years", "75+ years")) %>%
  ggplot(aes(x = age, y = suicides)) +
  geom_col(fill = "lightblue") +
  labs(y = "Suicides", x = "Age") +
  scale_y_continuous(breaks = c(500000, 1000000, 1500000, 2000000, 2500000), labels = c("50k", "100k", "150k", "200k", "250k"))
```

--

---

#What did we learn?

<ul>
<br>
--
<font size="13"><li>35-54 year olds seem to have the highest suicide count<br><br></font>

--
<font size="13"><li>Although this is the count so it does not take into play the rate.<br><br></font>

--
<font size="13"><li>The proportion of ages will be different.<br><br></font>
</ul>

---

#Suicide Hotline

--
<font size="13">Thank you for watching our presentation.<br><br></font>
--

<font size="13">Below is a number for a mental health hotline.<br><br></font>

<font size="13">samaritans: 116 123<br><br></font>

